Calibration tests beyond classification

### 2021-03-17 paper/main.tex
Move figures to corresponding sections to avoid excessive empty space
Fix typos

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[t]{0.62\textwidth}
        \centering
        \textbf{a}
        \includegraphics[valign=t,scale=0.23]{figures/ols/heatmap.pdf}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[t]{0.38\textwidth}
        \centering
        \textbf{b}
        \includegraphics[valign=t,scale=0.23]{figures/ols/quantiles.pdf}
    \end{subfigure}
    \caption{Illustration of a conditional distribution $\Prob(Y|X)$ with scalar
    feature and target. We consider a Gaussian predictive model $P$, obtained by
    ordinary least squares regression with 100 training data points (orange dots).
    Empirically the predicted quantiles on 50 validation data points appear close
    to being calibrated, although model $P$ is uncalibrated according to
    \cref{def:calibration}. Using the framework in this
    paper, on the same validation data a statistical test allows us to reject
    the null hypothesis that model $P$ is calibrated at a significance level of
    $\alpha = 0.05$ ($p < 0.05$). See~\cref{app:ols} for details.}%
    \label{fig:ols_motivation}
    %\vspace{-0.5\baselineskip}
\end{figure}

@@

Our framework can be used to reason about and test for calibration of \emph{any
probabilistic predictive model}. As explained above, this is in stark contrast with
existing methods
that are restricted to simple output distributions, such as classification
and \emph{scalar-valued} regression problems. 
A \emph{key contribution} of this paper is a new framework that
is applicable to \emph{multivariate} regression, as well as situations when the output is of a
different (e.g. discrete ordinal) or more complex (e.g, graph-structured) type, with clear practical implications.

@@

The construction of the product space $\mathcal{P} \times \mathcal{Y}$ suggests
the use of tensor product kernels $k = k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$, where
$k_{\mathcal{P}} \colon \mathcal{P} \times \mathcal{P} \to \mathbb{R}$ and
$k_{\mathcal{Y}} \colon \mathcal{Y} \times \mathcal{Y} \to \mathbb{R}$ are
kernels on the spaces of predicted distributions and targets, respectively.\footnote{As
mentioned above, our framework rephrases and generalizes the construction used
by \citet{Widmann2019}. The matrix-valued kernels that they employ can be
recovered by setting $k_{\mathcal{P}}$ to a Gaussian kernel on the probability simplex and
$k_{\mathcal{Y}}(y_1, y_2) = \delta_{y_1,y_2}$.}

@@

In machine learning, common
probabilistic predictive models output parameters of distributions such as
mean and variance of normal distributions. Naturally these
parameterizations give rise to injective mappings
$\phi \colon \mathcal{P} \to \mathbb{R}^d$ that can be used to define a
Hilbertian metric
\begin{equation*}\label{eq:hilbertian}
    d_{\mathcal{P}}(p, p') = {\|\phi(p) - \phi(p')\|}_2.
\end{equation*}
For such metrics, $k_{\mathcal{P}}$ in \cref{eq:kernelp}
is a valid kernel for all $\lambda > 0$ and
$\nu \in (0, 2]$~\citep[Corollary~3.3.3, Proposition~3.2.7]{Berg1984}.
In~\cref{app:mixture} we show that for many mixture models, and hence model ensembles,
Hilbertian metrics between model components
can be lifted to Hilbertian metrics between mixture models. This construction
is a generalization of the Wasserstein-like distance for Gaussian mixture models proposed
by \citet{Delon2019,Chen2019}.

@@

As an alternative to our construction based on the
joint distributions of $(P_X, Y)$ and $(P_X, Z_X)$, one could try
to directly compare the conditional distributions $\Prob(Y|P_X)$
and $\Prob(Z_X|P_X) = P_X$. For instance, \citet{Ren2016} proposed the
conditional MMD based on the so-called conditional kernel mean
embedding~\citep{Song2009,Song2013}. However, as noted by
\citet{Park2020}, its common definition as operator
between two RKHS is based on very restrictive assumptions, which are violated
in many situations~\citep[see, e.g.,][Footnote~4]{Fukumizu2013}
and typically require regularized estimates. Hence, even theoretically,
often the conditional MMD is
\textquote[\cite{Park2020}]{not an exact measure of discrepancy between conditional distributions}.
In contrast, the maximum conditional mean discrepancy~(MCMD) proposed in a
concurrent work by \citet{Park2020} is a random variable derived from much
weaker measure-theoretical assumptions.
The MCMD provides a local discrepancy conditional on random predictions
whereas KCE is a global real-valued summary of these local
discrepancies.\footnote{In our calibration setting, MMCD is
almost surely equal to
$
    \sup_{f \in \mathcal{F}_\mathcal{Y}}
    \big| \Exp_{Y|P_X}\big(f(Y)|P_X\big) - \Exp_{Z_X|P_X}\big(f(Z_X)|P_X\big) \big|,
$
where $\mathcal{F}_{\mathcal{Y}} \coloneqq \{ f \colon \mathcal{Y} \to \mathbb{R} | \|f\|_{\mathcal{H}_{\mathcal{Y}}} \leq 1 \}$
for an RKHS $\mathcal{H}_{\mathcal{Y}}$ with kernel
$k_{\mathcal{Y}} \colon \mathcal{Y} \times \mathcal{Y} \to \mathbb{R}$.
If kernel $k_{\mathcal{Y}}$ is characteristic, $\mathrm{MCMD} = 0$ almost
surely if and only if model $P$ is calibrated~\citep[Theorem~3.7]{Park2020}. Although the definition of MCMD
only requires
a kernel $k_{\mathcal{Y}}$ on the target space, a kernel $k_{\mathcal{P}}$ on the space of predictions
has to be specified for the evaluation of its regularized estimates.}

@@

Additionally, we evaluate empirical test errors of calibration tests
at a fixed significance level $\alpha = 0.05$. The evaluations, visualized in
\cref{fig:synthetic_tests_10_main} for models with ten-dimensional targets, demonstrate
empirically that the percentage of incorrect rejections of $H_0$ converges to the set
significance level as the number of samples increases.
Moreover, the results highlight the computational burden of the calibration test
that estimates quantiles of the intractable asymptotic distribution of
$\widehat{\mathrm{SKCE}}_{k,n}$ by bootstrapping. As expected, due to the larger
variance of $\widehat{\mathrm{SKCE}}_{k,2}$ the test with fixed block size $B = 2$
shows a decreased test power although being computationally much more efficient.

\begin{figure}[!hbtp]
    \centering
    \includegraphics[scale=0.35]{figures/synthetic/tests_dim=10.pdf}
    \caption{Empirical test errors for 500 data sets of $n \in \{4, 16, 64, 256, 1024\}$
    samples from models with targets of dimension $d = 10$. The dashed black line
    indicates the set signficance level $\alpha = 0.05$.}%
    \label{fig:synthetic_tests_10_main}
    \vspace{-\baselineskip}
\end{figure}

\begin{figure}[!hbtp]
    \centering
    \includegraphics[scale=0.35]{figures/friedman/statsplot_zoom.pdf}
    \caption{Mean squared error (MSE), average negative log-likelihood (NLL),
    $\widehat{\mathrm{SKCE}}_k$ (SKCE (biased)), and $p$-value approximation ($p$-value) of
    ten Gaussian predictive models for the Friedman~1 regression problem versus
    the number of training iterations.
    Evaluations on the training data set (100 samples) are displayed in green
    and orange, and on the test data set (50 samples) in blue and purple.
    The green and blue line and their surrounding bands represent
    the mean and the range of the evaluations of the ten models. The orange and
    purple lines visualize the evaluations of the ensemble models.}%
    \label{fig:friedman1_zoom}
    \vspace{-\baselineskip}
\end{figure}

@@

The Friedman~1 regression problem~\citep{Friedman1979,Friedman1983,Friedman1991}
is a classic non-linear regression problem with ten-dimensional features and
real-valued targets with Gaussian noise. We train a Gaussian predictive model
whose mean is modelled by a shallow neural network and %whose variance is a
a single scalar
variance
parameter (consistent with the data-generating model) ten times
with different initial parameters. \Cref{fig:friedman1_zoom} shows estimates of the
mean-squared error (MSE),
the average negative log-likelihood (NLL), $\mathrm{SKCE}_k$, and a $p$-value
approximation for these models and their ensemble on the training and a separate
test data set.
All estimates indicate consistently
that the models are overfit after 1500 training iterations. The estimations of
$\mathrm{SKCE}_k$ and the $p$-values allow to focus on calibration specifically,
whereas MSE indicates accuracy only and NLL, as any proper scoring rule~\citep{Broecker2009}, provides a summary of calibration and accuracy. The estimation
of $\mathrm{SKCE}_k$ in addition to NLL could serve as another source of information
for early stopping and model selection.

@@

We consider a regression with scalar feature $X$ and scalar target $Y$ with input-dependent Gaussian noise
that is inspired by a problem by \citet{Gustafsson2019}. Feature $X$ is
distributed uniformly at random in $[-1, 1]$, and target $Y$ is distributed according to
\begin{equation*}
    Y \sim \sin(\pi X) + | 1 + X | \epsilon,
\end{equation*}
where $\epsilon \sim \mathcal{N}(0, 0.15^2)$. We train a linear regression model $P$ with
homoscedastic variance using ordinary least squares and a data set of 100 i.i.d.\
pairs of feature $X$ and target $Y$ (see~\cref{fig:ols_heatmap}).

@@

A validation data set of $n = 50$ i.i.d.\ pairs of $X$ and $Y$ is used to
evaluate the empirical cumulative probability
\begin{equation*}
    n^{-1} \sum_{i=1}^n \mathbbm{1}_{[0,\tau]}\big(P(Y \leq Y_i | X = X_i)\big)
\end{equation*}
of model $P$ for quantile levels $\tau \in [0, 1]$. Model $P$ would be quantile
calibrated~\citep{Song2019}
\begin{equation*}
    \tau = \Prob_{X',Y'}\big(P(Y \leq Y' | X = X') \leq \tau \big)
\end{equation*}
for all $\tau \in [0,1]$, where $(X, Y)$ and $(X', Y')$ are independent identically
distributed pairs of random variables (see~\cref{fig:ols_quantiles}).

@@

Additionally, we compute a $p$-value estimate of the null hypothesis $H_0$ that model $P$
is calibrated using an estimation of the quantile of the asymptotic distribution
of $n \widehat{\mathrm{SKCE}}_{k,n}$ with 100000 bootstrap samples
on the validation data set (see~\cref{remark:skceb_quadratic}). Kernel $k$ is chosen
as a tensor product kernel of two Gaussian kernels
\begin{equation*}
    \begin{split}
        k\big((p, y), (p', y')\big) &= \exp{\big(-W_2(p, p')\big)} \exp{\big(-(y - y')^2/2 \big)} \\
        &= \exp{\bigg(-\sqrt{(m_p - m_{p'})^2 - (\sigma_p - \sigma_{p'})^2}\bigg)} \exp{\big(- (y - y')^2/2\big)},
    \end{split}
\end{equation*}
where $W_2$ is the 2-Wasserstein distance and $m_p, m_{p'}$ and $\sigma_p, \sigma_{p'}$ denote
the mean and the standard deviation of the normal distributions $p$ and $p'$ (see~\cref{app:normal}).
We obtain $p < 0.05$ in our experiment, and hence the calibration test rejects $H_0$ at the
significance level $\alpha = 0.05$.

@@

In the second setup we simulate an uncalibrated model of the form
\begin{equation*}
    Y \,|\, P_X = \mathcal{N}(\mu, \Sigma) \sim \mathcal{N}([0.1, \mu_2, \ldots, \mu_d], \Sigma).
\end{equation*}

We perform an evaluation of the convergence and computation time of the biased estimator
$\widehat{\mathrm{SKCE}}_k$ and the unbiased estimator $\widehat{\mathrm{SKCE}}_{k,B}$ with
blocks of size $B \in \{2, \sqrt{n}, n\}$. We use a tensor product kernel
\begin{equation*}
\begin{split}
    k\big((p, y), (p', y')\big) &= \exp{\big(- W_2(p, p')\big)} \exp{\big(-(y - y')^2/2\big)} \\
    &= \exp{\bigg(-\sqrt{(m_p - m_{p'})^2 - (\sigma_p - \sigma_{p'})^2}\bigg)} \exp{\big( - (y - y')^2/2\big)},
\end{split}
\end{equation*}
where $W_2$ is the 2-Wasserstein distance and $m_p, m_{p'}$ and $\sigma_p, \sigma_{p'}$
denote the mean and the standard deviation of the normal distributions $p$ and $p'$.

@@

We compare calibration tests based on the (tractable) asymptotic distribution of
$\widehat{\mathrm{SKCE}}_{k,B}$ with fixed block size $B \in \{2, \sqrt{n}$
(see~\cref{remark:skceb_fixed}), the (intractable) asymptotic distribution of
$\widehat{\mathrm{SKCE}}_{k,n}$ which is approximated with 1000 bootstrap
samples (see~\cref{remark:skceb_quadratic}), and a Hotelling's $T^2$-statistic for
$\mathrm{UCME}_k$ with 10 test locations (see~\cref{app:ucme}). We compute
the empirical test errors (percentage of false rejections of the null hypothesis $H_0$
that model $P$ is calibrated if $P$ is calibrated, and percentage of false non-rejections
of $H_0$ if $P$ is not calibrated) at a fixed significance level $\alpha = 0.05$ and
the minimal computation time for the calibrated and the uncalibrated model with
dimensions $d = 1$ and $d = 10$ for $500$ independently drawn data sets of
$n \in \{4, 16, 64, 256, 1024\}$ samples of $(P_X, Y)$. The 10 test
predictions of the $\mathrm{CME}$ test are of the form $\mathcal{N}(m, 0.1^2 \mathbf{I}_d)$
where $m$ is distributed uniformly at random in the $d$-dimensional unit hypercube $[0,1]^d$,
the corresponding 10 test targets are i.i.d.\ according to
$\mathcal{N}(\mathbf{0}, 0.1^2 \mathbf{I}_d)$.

\Cref{fig:synthetic_tests_1,fig:synthetic_tests_10} show that all tests adhere to the set
significance level asymptotically as the number of samples increases. The convergence of
the $\mathrm{CME}$ test with 10 test locations is found to be much slower than the convergence
of all other tests.
The tests based on tractable asymptotic distribution of $\widehat{\mathrm{SKCE}}_{k,B}$ for
fixed block size $B$ are orders of magnitudes faster than the test based on the intractable
asymptotic distribution of $\widehat{\mathrm{SKCE}}_{k,B}$, approximated with 1000 bootstrap
samples. We see that the efficiency gain comes at the cost of decreased test power for smaller
number of samples, explained by the increasing variance of $\widehat{\mathrm{SKCE}}_{k,B}$ for
decreasing block sizes $B$.
However, in our examples the test based on $\widehat{\mathrm{SKCE}}_{k,\sqrt{n}}$
still achieves good test power for reasonably large number of samples (> 30).

@@

We train 10 models with different initializations of parameters $\theta$. The initial values
of the weight matrices of the neural networks are sampled from the uniform Glorot
initialization~\citep{Glorot2010} and the offset vectors are initialized with zeros.
In \cref{fig:friedman1}, we visualize estimates of accuracy and calibration measures
on the training and test data set with 100 and 50 samples, respectively, for
5000 training iterations. The pinball loss is a common measure and
training objective for calibration of quantiles~\citep{Song2019}. It is defined
as
\begin{equation*}
    \Exp_{X,Y} L_{\tau}\big(Y, \mathrm{quantile}(P_X, \tau)\big),
\end{equation*}
where $L_{\tau}(y, \tilde{y}) = (1 - \tau) (\tilde{y} - y)_{+} + \tau (y - \tilde{y})_{+}$
and $\mathrm{quantile}(P_x, \tau) = \inf_{y} \{P_x(Y \leq y) \geq \tau\}$ for
quantile level $\tau \in [0, 1]$.
In \cref{fig:friedman1} we plot the average pinball loss (pinball) for
quantile levels $\tau \in \{0.05, 0.1, \ldots, 0.95\}$. Estimates of the $\mathrm{SKCE}_{k,n}$
(SKCE (unbiased)) and $\mathrm{SKCE}_{k}$ (SKCE (biased)) are evaluated for the tensor product kernel
\begin{equation*}
    \begin{split}
        k\big((p, y), (p', y')\big) &= \exp{\big(-W_2(p, p')\big)} \exp{\big(-(y - y')^2/2 \big)} \\
        &= \exp{\bigg(-\sqrt{(m_p - m_{p'})^2 - (\sigma_p - \sigma_{p'})^2}\bigg)} \exp{\big(- (y - y')^2/2\big)},
    \end{split}
\end{equation*}
where $W_2$ is the 2-Wasserstein distance and $m_p, m_{p'}$ and $\sigma_p, \sigma_{p'}$ denote
the mean and the standard deviation of the normal distributions $p$ and $p'$ (see~\cref{app:normal}).
The $p$-value estimate ($p$-value) is computed by estimating the quantile of the asymptotic distribution
of $n \widehat{\mathrm{SKCE}}_{k,n}$ with 1000 bootstrap samples
(see~\cref{remark:skceb_quadratic}). The estimates of the mean squared error
and the average negative log-likelihood are denoted by MSE and NLL.
All estimators indicate consistently that the trained models suffer from overfitting
after around 1000 training iterations.

Additionally, we form ensemble models of the ten individual models at every training
iteration. The evaluations for the ensembles are visualized in \cref{fig:friedman1} as
well. Apart from the unbiased estimates of $\mathrm{SKCE}_k$, the estimates of the
ensemble models are consistently better than the average estimates of the ensemble
members. For the mean squared error and the negative log-likelihood this behaviour
is guaranteed theoretically by the generalized mean inequality.

>>>>

\begin{figure}[!htb]
    \centering
    \begin{subfigure}[t]{0.62\textwidth}
        \centering
        \textbf{a}
        \includegraphics[valign=t,scale=0.23]{figures/ols/heatmap.pdf}
    \end{subfigure}%
    \hfill%
    \begin{subfigure}[t]{0.38\textwidth}
        \centering
        \textbf{b}
        \includegraphics[valign=t,scale=0.23]{figures/ols/quantiles.pdf}
    \end{subfigure}
    \caption{Illustration of a conditional distribution $\Prob(Y|X)$ with scalar
    feature and target. We consider a Gaussian predictive model $P$, obtained by
    ordinary least squares regression with 100 training data points (orange dots).
    Empirically the predicted quantiles on 50 validation data points appear close
    to being calibrated, although model $P$ is uncalibrated according to
    \cref{def:calibration}. Using the framework in this
    paper, on the same validation data a statistical test allows us to reject
    the null hypothesis that model $P$ is calibrated at a significance level of
    $\alpha = 0.05$ ($p < 0.05$). See~\cref{app:ols} for details.}%
    \label{fig:ols_motivation}
\end{figure}

@@

Our framework can be used to reason about and test for calibration of \emph{any
probabilistic predictive model}. As explained above, this is in stark contrast with
existing methods
that are restricted to simple output distributions, such as classification
and \emph{scalar-valued} regression problems. 
A \emph{key contribution} of this paper is a new framework that
is applicable to \emph{multivariate} regression, as well as situations when the output is of a
different (e.g., discrete ordinal) or more complex (e.g., graph-structured) type, with clear practical implications.

@@

The construction of the product space $\mathcal{P} \times \mathcal{Y}$ suggests
the use of tensor product kernels $k = k_{\mathcal{P}} \otimes k_{\mathcal{Y}}$, where
$k_{\mathcal{P}} \colon \mathcal{P} \times \mathcal{P} \to \mathbb{R}$ and
$k_{\mathcal{Y}} \colon \mathcal{Y} \times \mathcal{Y} \to \mathbb{R}$ are
kernels on the spaces of predicted distributions and targets, respectively.\footnote{As
mentioned above, our framework rephrases and generalizes the construction used
by \citet{Widmann2019}. The matrix-valued kernels that they employ can be
recovered by setting $k_{\mathcal{P}}$ to a Laplacian kernel on the probability simplex and
$k_{\mathcal{Y}}(y, y') = \delta_{y,y'}$.}

@@

In machine learning, common
probabilistic predictive models output parameters of distributions such as
mean and variance of normal distributions. Naturally these
parameterizations give rise to injective mappings
$\phi \colon \mathcal{P} \to \mathbb{R}^d$ that can be used to define a
Hilbertian metric
\begin{equation*}\label{eq:hilbertian}
    d_{\mathcal{P}}(p, p') = {\|\phi(p) - \phi(p')\|}_2.
\end{equation*}
For such metrics, $k_{\mathcal{P}}$ in \cref{eq:kernelp}
is a valid kernel for all $\lambda > 0$ and
$\nu \in (0, 2]$~\citep[Corollary~3.3.3, Proposition~3.2.7]{Berg1984}.
In~\cref{app:mixture} we show that for many mixture models, and hence model ensembles,
Hilbertian metrics between model components
can be lifted to Hilbertian metrics between mixture models. This construction
is a generalization of the Wasserstein-like distance for Gaussian mixture models proposed
by \citet{Delon2019,Chen2019,Chen2020}.

@@

As an alternative to our construction based on the
joint distributions of $(P_X, Y)$ and $(P_X, Z_X)$, one could try
to directly compare the conditional distributions $\Prob(Y|P_X)$
and $\Prob(Z_X|P_X) = P_X$. For instance, \citet{Ren2016} proposed the
conditional MMD based on the so-called conditional kernel mean
embedding~\citep{Song2009,Song2013}. However, as noted by
\citet{Park2020}, its common definition as operator
between two RKHS is based on very restrictive assumptions, which are violated
in many situations~\citep[see, e.g.,][Footnote~4]{Fukumizu2013}
and typically require regularized estimates. Hence, even theoretically,
often the conditional MMD is
\textquote[\cite{Park2020}]{not an exact measure of discrepancy between conditional distributions}.
In contrast, the maximum conditional mean discrepancy~(MCMD) proposed in a
concurrent work by \citet{Park2020} is a random variable derived from much
weaker measure-theoretical assumptions.
The MCMD provides a local discrepancy conditional on random predictions
whereas KCE is a global real-valued summary of these local
discrepancies.\footnote{In our calibration setting, the $\mathrm{MCMD}$ is
almost surely equal to
$
    \sup_{f \in \mathcal{F}_\mathcal{Y}}
    \big| \Exp_{Y|P_X}\big(f(Y)|P_X\big) - \Exp_{Z_X|P_X}\big(f(Z_X)|P_X\big) \big|,
$
where $\mathcal{F}_{\mathcal{Y}} \coloneqq \{ f \colon \mathcal{Y} \to \mathbb{R} | \|f\|_{\mathcal{H}_{\mathcal{Y}}} \leq 1 \}$
for an RKHS $\mathcal{H}_{\mathcal{Y}}$ with kernel
$k_{\mathcal{Y}} \colon \mathcal{Y} \times \mathcal{Y} \to \mathbb{R}$.
If kernel $k_{\mathcal{Y}}$ is characteristic, $\mathrm{MCMD} = 0$ almost
surely if and only if model $P$ is calibrated~\citep[Theorem~3.7]{Park2020}. Although the definition of MCMD
only requires
a kernel $k_{\mathcal{Y}}$ on the target space, a kernel $k_{\mathcal{P}}$ on the space of predictions
has to be specified for the evaluation of its regularized estimates.}

@@

Additionally, we evaluate empirical test errors of calibration tests
at a fixed significance level $\alpha = 0.05$. The evaluations, visualized in
\cref{fig:synthetic_tests_10_main} for models with ten-dimensional targets, demonstrate
empirically that the percentage of incorrect rejections of $H_0$ converges to the set
significance level as the number of samples increases.
Moreover, the results highlight the computational burden of the calibration test
that estimates quantiles of the intractable asymptotic distribution of
$n\widehat{\mathrm{SKCE}}_{k,n}$ by bootstrapping. As expected, due to the larger
variance of $\widehat{\mathrm{SKCE}}_{k,2}$ the test with fixed block size $B = 2$
shows a decreased test power although being computationally much more efficient.

\begin{figure}[!hbtp]
    \centering
    \includegraphics[scale=0.35]{figures/synthetic/tests_dim=10.pdf}
    \caption{Empirical test errors for 500 data sets of $n \in \{4, 16, 64, 256, 1024\}$
    samples from models with targets of dimension $d = 10$. The dashed black line
    indicates the set signficance level $\alpha = 0.05$.}%
    \label{fig:synthetic_tests_10_main}
\end{figure}

@@

The Friedman~1 regression problem~\citep{Friedman1979,Friedman1983,Friedman1991}
is a classic non-linear regression problem with ten-dimensional features and
real-valued targets with Gaussian noise. We train a Gaussian predictive model
whose mean is modelled by a shallow neural network and
a single scalar
variance
parameter (consistent with the data-generating model) ten times
with different initial parameters. \Cref{fig:friedman1_zoom} shows estimates of the
mean squared error (MSE),
the average negative log-likelihood (NLL), $\mathrm{SKCE}_k$, and a $p$-value
approximation for these models and their ensemble on the training and a separate
test data set.
All estimates indicate consistently
that the models are overfit after 1500 training iterations. The estimations of
$\mathrm{SKCE}_k$ and the $p$-values allow to focus on calibration specifically,
whereas MSE indicates accuracy only and NLL, as any proper scoring rule~\citep{Broecker2009}, provides a summary of calibration and accuracy. The estimation
of $\mathrm{SKCE}_k$ in addition to NLL could serve as another source of information
for early stopping and model selection.

\begin{figure}[!hbtp]
    \centering
    \includegraphics[scale=0.35]{figures/friedman/statsplot_zoom.pdf}
    \caption{Mean squared error (MSE), average negative log-likelihood (NLL),
    $\widehat{\mathrm{SKCE}}_k$ (SKCE (biased)), and $p$-value approximation ($p$-value) of
    ten Gaussian predictive models for the Friedman~1 regression problem versus
    the number of training iterations.
    Evaluations on the training data set (100 samples) are displayed in green
    and orange, and on the test data set (50 samples) in blue and purple.
    The green and blue line and their surrounding bands represent
    the mean and the range of the evaluations of the ten models. The orange and
    purple lines visualize the evaluations of their ensemble.}%
    \label{fig:friedman1_zoom}
\end{figure}

@@

We consider a regression problem with scalar feature $X$ and scalar target $Y$ with input-dependent Gaussian noise
that is inspired by a problem by \citet{Gustafsson2019}. Feature $X$ is
distributed uniformly at random in $[-1, 1]$, and target $Y$ is distributed according to
\begin{equation*}
    Y \sim \sin(\pi X) + | 1 + X | \epsilon,
\end{equation*}
where $\epsilon \sim \mathcal{N}(0, 0.15^2)$. We train a linear regression model $P$ with
homoscedastic variance using ordinary least squares and a data set of 100 i.i.d.\
pairs of feature $X$ and target $Y$ (see~\cref{fig:ols_heatmap}).

@@

A validation data set of $n = 50$ i.i.d.\ pairs of $X$ and $Y$ is used to
evaluate the empirical cumulative probability
\begin{equation*}
    n^{-1} \sum_{i=1}^n \mathbbm{1}_{[0,\tau]}\big(P(Y \leq Y_i | X = X_i)\big)
\end{equation*}
of model $P$ for quantile levels $\tau \in [0, 1]$. Model $P$ would be quantile
calibrated~\citep{Song2019} if
\begin{equation*}
    \tau = \Prob_{X',Y'}\big(P(Y \leq Y' | X = X') \leq \tau \big)
\end{equation*}
for all $\tau \in [0,1]$, where $(X, Y)$ and $(X', Y')$ are independent identically
distributed pairs of random variables (see~\cref{fig:ols_quantiles}).

@@

Additionally, we compute a $p$-value estimate of the null hypothesis $H_0$ that model $P$
is calibrated using an estimation of the quantile of the asymptotic distribution
of $n \widehat{\mathrm{SKCE}}_{k,n}$ with 100000 bootstrap samples
on the validation data set (see~\cref{remark:skceb_quadratic}). Kernel $k$ is chosen
as the tensor product kernel
\begin{equation*}
    \begin{split}
        k\big((p, y), (p', y')\big) &= \exp{\big(-W_2(p, p')\big)} \exp{\big(-(y - y')^2/2 \big)} \\
        &= \exp{\bigg(-\sqrt{(m_p - m_{p'})^2 + (\sigma_p - \sigma_{p'})^2}\bigg)} \exp{\big(- (y - y')^2/2\big)},
    \end{split}
\end{equation*}
where $W_2$ is the 2-Wasserstein distance and $m_p, m_{p'}$ and $\sigma_p, \sigma_{p'}$ denote
the mean and the standard deviation of the normal distributions $p$ and $p'$ (see~\cref{app:normal}).
We obtain $p < 0.05$ in our experiment, and hence the calibration test rejects $H_0$ at the
significance level $\alpha = 0.05$.

@@

In the second setup we simulate an uncalibrated model of the form
\begin{equation*}
    Y \,|\, P_X = \mathcal{N}(\mu, \Sigma) \sim \mathcal{N}([0.1, \mu_2, \ldots, \mu_d]^{\mathsf{T}}, \Sigma).
\end{equation*}

We perform an evaluation of the convergence and computation time of the biased estimator
$\widehat{\mathrm{SKCE}}_k$ and the unbiased estimator $\widehat{\mathrm{SKCE}}_{k,B}$ with
blocks of size $B \in \{2, \sqrt{n}, n\}$. We use the tensor product kernel
\begin{equation*}
\begin{split}
    k\big((p, y), (p', y')\big) &= \exp{\big(- W_2(p, p')\big)} \exp{\big(-(y - y')^2/2\big)} \\
    &= \exp{\bigg(-\sqrt{(m_p - m_{p'})^2 + (\sigma_p - \sigma_{p'})^2}\bigg)} \exp{\big( - (y - y')^2/2\big)},
\end{split}
\end{equation*}
where $W_2$ is the 2-Wasserstein distance and $m_p, m_{p'}$ and $\sigma_p, \sigma_{p'}$
denote the mean and the standard deviation of the normal distributions $p$ and $p'$.

@@

We compare calibration tests based on the (tractable) asymptotic distribution of
$\sqrt{\lfloor n / B \rfloor} \widehat{\mathrm{SKCE}}_{k,B}$ with fixed block
size $B \in \{2, \sqrt{n}\}$
(see~\cref{remark:skceb_fixed}), the (intractable) asymptotic distribution of
$n\widehat{\mathrm{SKCE}}_{k,n}$ which is approximated with 1000 bootstrap
samples (see~\cref{remark:skceb_quadratic}), and a Hotelling's $T^2$-statistic for
$\mathrm{UCME}_{k,10}$ with 10 test locations (see~\cref{app:ucme}). We compute
the empirical test errors (percentage of false rejections of the null hypothesis $H_0$
that model $P$ is calibrated if $P$ is calibrated, and percentage of false non-rejections
of $H_0$ if $P$ is not calibrated) at a fixed significance level $\alpha = 0.05$ and
the minimal computation time for the calibrated and the uncalibrated model with
dimensions $d = 1$ and $d = 10$ for $500$ independently drawn data sets of
$n \in \{4, 16, 64, 256, 1024\}$ samples of $(P_X, Y)$. The 10 test
predictions of the $\mathrm{CME}$ test are of the form $\mathcal{N}(m, 0.1^2 \mathbf{I}_d)$
where $m$ is distributed uniformly at random in the $d$-dimensional unit hypercube $[0,1]^d$,
the corresponding 10 test targets are i.i.d.\ according to
$\mathcal{N}(\mathbf{0}, 0.1^2 \mathbf{I}_d)$.

\Cref{fig:synthetic_tests_1,fig:synthetic_tests_10} show that all tests adhere to the set
significance level asymptotically as the number of samples increases. The convergence of
the $\mathrm{CME}$ test with 10 test locations is found to be much slower than the convergence
of all other tests.
The tests based on the tractable asymptotic distribution of
$\sqrt{\lfloor n / B \rfloor} \widehat{\mathrm{SKCE}}_{k,B}$ for
fixed block size $B$ are orders of magnitudes faster than the test based on the intractable
asymptotic distribution of $n \widehat{\mathrm{SKCE}}_{k,n}$, approximated with 1000 bootstrap
samples. We see that the efficiency gain comes at the cost of decreased test power for smaller
number of samples, explained by the increasing variance of $\widehat{\mathrm{SKCE}}_{k,B}$ for
decreasing block sizes $B$.
However, in our examples the test based on $\widehat{\mathrm{SKCE}}_{k,\sqrt{n}}$
still achieves good test power for reasonably large number of samples (> 30).

@@

We train 10 models with different initializations of parameters $\theta$. The initial values
of the weight matrices of the neural networks are sampled from the uniform Glorot
initialization~\citep{Glorot2010} and the offset vectors are initialized with zeros.
In \cref{fig:friedman1}, we visualize estimates of accuracy and calibration measures
on the training and test data set with 100 and 50 samples, respectively, for
5000 training iterations. The pinball loss is a common measure and
training objective for calibration of quantiles~\citep{Song2019}. It is defined
as
\begin{equation*}
    \Exp_{X,Y} L_{\tau}\big(Y, \mathrm{quantile}(P_X, \tau)\big),
\end{equation*}
where $L_{\tau}(y, \tilde{y}) = (1 - \tau) (\tilde{y} - y)_{+} + \tau (y - \tilde{y})_{+}$
and $\mathrm{quantile}(P_x, \tau) = \inf_{y} \{P_x(Y \leq y) \geq \tau\}$ for
quantile level $\tau \in [0, 1]$.
In \cref{fig:friedman1} we plot the average pinball loss (pinball) for
quantile levels $\tau \in \{0.05, 0.1, \ldots, 0.95\}$. We evaluate
$\widehat{\mathrm{SKCE}}_{k,n}$ (SKCE (unbiased)) and $\widehat{\mathrm{SKCE}}_{k}$
(SKCE (biased)) for the tensor product kernel
\begin{equation*}
    \begin{split}
        k\big((p, y), (p', y')\big) &= \exp{\big(-W_2(p, p')\big)} \exp{\big(-(y - y')^2/2 \big)} \\
        &= \exp{\bigg(-\sqrt{(m_p - m_{p'})^2 + (\sigma_p - \sigma_{p'})^2}\bigg)} \exp{\big(- (y - y')^2/2\big)},
    \end{split}
\end{equation*}
where $W_2$ is the 2-Wasserstein distance and $m_p, m_{p'}$ and $\sigma_p, \sigma_{p'}$ denote
the mean and the standard deviation of the normal distributions $p$ and $p'$ (see~\cref{app:normal}).
The $p$-value estimate ($p$-value) is computed by estimating the quantile of the asymptotic distribution
of $n \widehat{\mathrm{SKCE}}_{k,n}$ with 1000 bootstrap samples
(see~\cref{remark:skceb_quadratic}). The estimates of the mean squared error
and the average negative log-likelihood are denoted by MSE and NLL.
All estimators indicate consistently that the trained models suffer from overfitting
after around 1000 training iterations.

Additionally, we form ensembles of the ten individual models at every training
iteration. The evaluations for the ensembles are visualized in \cref{fig:friedman1} as
well. Apart from the unbiased estimates of $\mathrm{SKCE}_k$, the estimates of the
ensembles are consistently better than the average estimates of the ensemble
members. For the mean squared error and the negative log-likelihood this behaviour
is guaranteed theoretically by the generalized mean inequality.

\begin{figure}[htp]
    \begin{center}
        \includegraphics[scale=0.35]{figures/friedman/statsplot.pdf}
        \caption{Estimates of different accuracy and calibration measures of ten
        Gaussian predictive models for the Friedman~1 regression problem versus
        the number of training iterations.
        Evaluations on the training data set (100 samples) are displayed in green
        and orange, and on the test data set (50 samples) in blue and purple.
        The green and blue line and their surrounding bands represent
        the mean and the range of the evaluations of the ten models. The orange and
        purple lines visualize the evaluations of their ensemble.}
        \label{fig:friedman1}
    \end{center}
\end{figure}
